Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-02-06 analysis/paper/paper.Rmd
fix typos

---
title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan. We infer increasing craft specialization and changes in social organization. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. This study implies that craft specialization can be a deliberate act of resistance to show indigenous identity and ethnicity in a direct culture contact situation.
keywords: |
  Geometric Morphometrics; ceramic specialization; Iron age; culture contact; standardization
highlights: |
  These are the highlights. 
---

@@

The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies. The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017]. @Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation. Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies. Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies. Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century. Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan. This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary. Here we analyze pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardization of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

@@

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation. Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standarization from different periods. Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

@@

The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan. In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across  the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods. We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The most abundant artefacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site. More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed. There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between. Those pots forms show high consistency in shape. They have a globular body with a short neck and wide mouth. The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs. These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors. Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests these globular pots were mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.

@@

A total of `r total_no` largely intact vessels from pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) were selected for elliptic Fourier analysis. Most of them cover every vessel component from rim to bottom. The shape of vessel is the major object analyzed in this paper, however, to fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r post_e`, Chinese = `r chi`) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. In addition, we use Monte Carlo method to test for randomness in spatial locations of pots to explore whether the distribution is meaningful indicated by clustered or dispersed pattern. As craft specialization increases, pottery distribution might present a shift from dispersed pattern to concentrated clusters that hints the emergence of specialized groups.

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan. All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually. Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel. Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases. The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .

@@

To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014]. While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].

@@

To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The first PC, vessel height, shows higher variation in the pre-European period compared to the Chinese period. That indicates that shape standardization was higher in the Chinese period compared to the pre-European period. The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases. To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs. P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European(Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)). A significant difference is also detected between the pre-European and post-European periods.

@@

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at the emergence of social inequality in the Indigenous community. Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999]. Our results indicate that differences in pottery shape can be detected using EFA. The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence. These changes even more pronounced during the Chinese contact period. In general, pottery vessels become shorter in height over time, that leading to an oval-shaped body.

Our CV tests indicates that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offers limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)). Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence. After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05). Although these results shows the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantially way reflected by overall shapes of pottery, not only limited to diameter variables. In addition, results of spatial analysis for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence. However, the distribution of pottery is more similar to the random distributions during the Chinese period. This contradicts to our expectation that clustered patten will be observed with an increase in pottery standardization since the emergence of specialized groups. Or, this indicates that there are other factors lead to pottery standardization instead of a small number of artisans.

>>>>

---
title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan. We infer increasing craft specialization and changes in social organization. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. This study implies that craft specialization can be a deliberate act of resistance to show indigenous identity and ethnicity in a direct culture contact situation.
keywords: |
  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization
highlights: |
  These are the highlights. 
---

@@

The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of the world, for example, when foreign trade goods are introduced to local Indigenous societies. The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017]. @Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation. Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies. Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies. Previous reports of archaeological evidence in northeastern Taiwan describes increases in the use of prestige goods, such as trade ornaments, and uneven artifact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century. Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan. This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is sensitive to non-linear aspects of shape variation is necessary. Here we analyze pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by increased standardization of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by greater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

@@

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contain crucial variation. Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottles excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods. Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

@@

The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan. In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods. We identified these phases using 32 previously published radiocarbon ages [@Chen2007], excavation depth measurements, stratigraphic details reported by the excavators (color, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The most abundant artifacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site. More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed. There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between. Those pots forms show high consistency in shape. They have a globular body with a short neck and wide mouth. The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs. These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors. Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests these globular pots were mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.

@@

A total of `r total_no` largely intact vessels from pre-contact contexts (n = `r pre_no`), post-European contexts (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`) were selected for elliptic Fourier analysis. Most of them cover every vessel component from rim to bottom. The shape of vessel is the major object analyzed in this paper, however, to fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of `r metric_all` reconstructed pots (pre-European = `r pre`, European = `r post_e`, Chinese = `r chi`) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were taken to compute an average value. To obtain information about the pottery shape, we compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. In addition, we use Monte Carlo method to test for randomness in spatial locations of pots to explore whether the distribution is meaningful indicated by clustered or dispersed pattern. As craft specialization increases, pottery distribution might present a shift from dispersed pattern to concentrated clusters that hints the emergence of specialized groups.

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan. All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually. Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel. Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyze shapes, especially for outline analysis [@Bonhomme2014]. The digilized outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases. The harmonic coefficients generated by EFA were analyzed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

To answer the question of whether CV values for vessel samples across our three occupational phases are significantly different or not, we used the Modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014]. While previous work has used the @Feltz1996 asymptotic test for the equality of coefficients of variation from k populations [@Lycett2008; @Okumura2014; @Eerkens2000; @Eerkens2001; @Hoggard2017], we prefer the MSLR test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].

@@

To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The first PC, vessel height, shows higher variation in the pre-European period compared to the Chinese period. That indicates that shape standardization was higher in the Chinese period compared to the pre-European period. The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases. To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs. P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)). A significant difference is also detected between the pre-European and post-European periods.

@@

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at the emergence of social inequality in the Indigenous community. Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999]. Our results indicate that differences in pottery shape can be detected using EFA. The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence. These changes even more pronounced during the Chinese contact period. In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)). Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence. After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05). Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables. In addition, results of spatial analysis for pottery shape samples show a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence. However, the distribution of pottery is more similar to the random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. Or, this indicates that there are other factors lead to pottery standardization instead of a small number of artisans.
